---
knitr:
  opts_chunk:
    cache.path: "../_cache/cls-neural-nets"
---

# Neural Network Classifiers {#sec-cls-neural-nets}

```{r}
#| label: cls-nnet-knitr-setup
#| include: false
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  fig.align = 'center',
  fig.width = 10,
  fig.height = 6,
  out.width = "95%",
  dev = 'svg',
  dev.args = list(bg = "transparent"),
  tidy = FALSE,
  echo = TRUE
)

options(digits = 4, width = 84)
options(dplyr.print_min = 6, dplyr.print_max = 6)
options(cli.width = 85)
options(crayon.enabled = FALSE)
options(pillar.advice = FALSE, pillar.min_title_chars = Inf, pillar.sigfig = 4)

source("../R/_common.R")
# fmt: skip
req_pkg <- 
  c("bestNormalize", "brulee", "embed", "mirai", "nnet", "probably",
    "spatialsample", "tidymodels")

req_pkg_fmt <- purrr::map_chr(req_pkg, ~ pkg_chr(.x))
```

This chapter describes how to use neural networks. 

## Requirements

You'll need `r length(req_pkg)` packages (`r req_pkg_fmt`) for this chapter:

```{r}
#| label: cls-nnet-installs
#| eval: false
#| echo: true
# fmt: skip
req_pkg <- 
  c("bestNormalize", "brulee", "embed", "mirai", "nnet", "probably",
    "spatialsample", "tidymodels")

# Check to see if they are installed:
pkg_installed <- vapply(req_pkg, rlang::is_installed, logical(1))

# Install missing packages:
if (any(!pkg_installed)) {
  install_list <- names(pkg_installed)[!pkg_installed]
  pak::pak(install_list)
}
```

Let's load the meta package and manage some between-package function conflicts. 

```{r}
#| label: start-tidymodels
#| results: hide
#| message: false
#| warning: false
library(tidymodels)
library(spatialsample)
library(embed)
library(bestNormalize)
library(discrim)
library(important)
library(probably)
library(mirai)

tidymodels_prefer()
theme_set(theme_bw() + theme(legend.position = "top"))
daemons(parallel::detectCores())

# Some general settings for tuning classification models:
cls_mtr <- metric_set(brier_class, roc_auc, pr_auc, mn_log_loss)
ctrl_grid <-
  control_grid(
    save_pred = TRUE,
    save_workflow = TRUE,
    parallel_over = "everything"
  )
```

As before, we'll load the objects that were already computed for the forestation data: 

```{r}
#| label: data-load
# "https://raw.githubusercontent.com/aml4td/website/main/RData/forested_data.RData" |>
#   url() |>
#   load()
load("~/content/website/RData/forested_data.RData")
```

## Neural Networks via Multilayer Perceptrons {#sec-cls-nnet}

There are several engines in tidymodels for fitting basic, feed-forward neural networks for classification: `nnet`, `keras`, `brulee`, and `brulee_two_layer`. The first three assume a single hidden layer. The focus here will be on using the `brulee` engine, as it offers the most options.

The `r pkg(parsnip)` package function for this model is `mlp()` and has main arguments: 

- `hidden_units`: the number of hidden units in the model.
- `activation`: the activation function (e.g., `"relu"`).
- `penalty`: the amount of penalization. 
- `dropout`: the proportion of model coefficients (i.e., weights) to set to zero during optimization. 
- `epochs`: the number of full passes through the training set.
- `learn_rate`: the learning rate. 

Please note that not all engines support all of these parameters. For example, the `"net"` engine cannot use `epoch`, `dropout`, `learn_rate`, or `activation`. 

For `brulee`, there are a few helpful engine-specific arguments: 

- `momentum`: the number used to specify historical gradient information during optimization.
- `batch_size`: an integer for the number of training set points in each batch.
- `class_weights`: numeric class weights. See `brulee::brulee_mlp()` for more details.
- `stop_iter`: a non-negative integer for how many iterations with no improvement before stopping. (default: 5L).
- `rate_schedule`: A function to change the learning rate over epochs. See `brulee::schedule_decay_time()` for details. Possible values include: `"decay_time"`, `"decay_expo"`, `"none"`, `"cyclic"`, and `"step"`.
- `optimizer`: the optimization method. Some example values are: `"SGD"`, `"ADAMw"`, `"Adadelta"`, `"Adagrad"`, and `"RMSprop"`. 

To demonstrate, we'll train our model using the AdamW variant of stochastic gradient descent for a maximum of 25 epochs, but may stop early after five consecutive poor iterations. 

Before fitting the model, some preprocessing is necessary. We will need to convert `county` into one or more numeric columns. We can use binary indicators; however, for demonstration purposes, a supervised effect encoding model is employed to generate a single column of values. After this, we use the orderNorm technique to transform our predictors to have the same standard normal distribution. 

```{r}
#| label: mlp-preprocess
encode_rec <-
  recipe(class ~ ., data = forested_train) |>
  step_lencode_mixed(county, outcome = "class") |>
  step_orderNorm(all_numeric_predictors())
```

Our model specification will tune numrous parameters: 

```{r}
#| label: mlp-spec
mlp_spec <-
  mlp(
    hidden_units = tune(),
    penalty = tune(),
    learn_rate = tune(),
    epochs = 25,
    activation = tune()
  ) |>
  set_engine(
    "brulee",
    stop_iter = 5,
    optimizer = "ADAMw",
    rate_schedule = tune(),
    batch_size = tune(),
    momentum = tune()
  ) |>
  set_mode("classification")

mlp_flow <- workflow(encode_rec, mlp_spec)
```

tidymodels is well-versed in each of these parameters and can default to sensible ranges for each. However, we will adjust a few parameter ranges and reduce the number of possible values for the learning rate scheduler and the activation function (for demonstration purposes). 

```{r}
#| label: mlp-param
mlp_param <-
  mlp_flow |>
  extract_parameter_set_dials() |>
  update(
    momentum = momentum(c(0.8, 0.99)),
    penalty = penalty(c(-10, -1)),
    learn_rate = learn_rate(c(-4, -1)),
    activation = activation(c("elu", "relu", "tanh", "tanhshrink")),
    rate_schedule = rate_schedule(c("cyclic", "decay_time", "none"))
  )
```

From here, we can use `tune_grid()` to run the computations. In the main text, several preprocessors were used with the same model specification. There, we used a workflow set similar to what was shown in the previous chapter for logistic regression. 

A note about `brulee`: it uses the `torch` ecosystem for computations natively in R (no Python required). We can run in parallel across CPU cores and/or, if your hardware allows, increase computational efficiency by utilizing a GPU.  

Note that it can be very difficult to write producible code with deep learning infrastructures like `torch` or `tensorflow`. When a GPU is used, it may be impossible to get exactly the same numbers from run to run. 

A space-filling design is used to tune: 

```{r}
#| label: mlp-tune
#| cache: true
set.seed(872)
forest_mlp_res <-
  mlp_flow |>
  tune_grid(
    resamples = forested_rs,
    grid = 25,
    param_info = mlp_param,
    metrics = cls_mtr,
    control = ctrl_grid
  )
```

For previous models, we use the `autoplot()` methods to quickly assess how well the model performed, identify which parameters were most important, and determine whether our grid search was searching the correct space. There is a constraint on `autoplot()` that prevents it from running with more than one qualitative parameter. Here we have two (`activation` and `rate_schedule`). 

We can write some custom code to approximate what we would get from `autoplot()`: 

```{r}
#| label: mlp-autoplot-ish
#| out-width: 100%
#| fig-width: 9
#| fig-height: 3.5
forest_mlp_res |>
  collect_metrics() |>
  filter(.metric == "brier_class") |>
  select(-n, -std_err) |>
  mutate(
    log_penalty = log10(penalty),
    log_learn_rate = log10(learn_rate),
    log_batch_size = log2(batch_size),
    .keep = "unused"
  ) |>
  pivot_longer(
    cols = c(hidden_units, starts_with("log_"), momentum),
    names_to = "Parameter",
    values_to = "Value"
  ) |>
  ggplot(aes(x = Value, y = mean, col = activation, pch = rate_schedule)) +
  geom_point() +
  facet_wrap(~Parameter, scales = "free_x", nrow = 1) +
  theme(legend.position = "top", legend.box = "vertical") +
  labs(y = "Brier Score")
```

`r back("cls-neural-nets.html#sec-cls-nnet")`

## Special Tabular Network Models {#sec-cls-tab-net}

### Batch Normalization {#sec-cls-batchnorm}

### Residual Networks {#sec-cls-resnet}

`r back("cls-neural-nets.html#sec-cls-resnet")`

### Attention Mechanisms {#sec-cls-attention}

`r back("cls-neural-nets.html#sec-cls-attention")`

### Multifocal Attention {#sec-cls-multi-attention}

`r back("cls-neural-nets.html#sec-cls-multi-attention")`

### Masking Mechanisms {#sec-cls-masking}

`r back("cls-neural-nets.html#sec-cls-masking")`

## Foundational Models {#sec-cls-foundational}

`r back("cls-neural-nets.html#sec-cls-foundational")`
