---
knitr:
  opts_chunk:
    cache.path: "../_cache/cls-trees/"
---

# Classification using Trees and Rules {#sec-cls-trees}

```{r}
#| label: cls-tree-knitr-setup
#| include: false

knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  fig.align = 'center',
  fig.width = 10,
  fig.height = 6,
  out.width = "95%",
  dev = 'svg',
  dev.args = list(bg = "transparent"),
  tidy = FALSE,
  echo = TRUE
)

options(digits = 4, width = 84)
options(dplyr.print_min = 6, dplyr.print_max = 6)
options(cli.width = 85)
options(crayon.enabled = FALSE)
options(pillar.advice = FALSE, pillar.min_title_chars = Inf, pillar.sigfig = 4)

source("../R/_common.R")
# fmt: skip
req_pkg <- 
  c("aorsf", "bonsai", "evtree", "partykit", "probably", "rpart", 
    "rules", "spatialsample", "tidymodels")

req_pkg_fmt <- purrr::map_chr(req_pkg, ~ pkg_chr(.x))
```

This chapter describes how to use neural networks. 

## Requirements

You'll need `r length(req_pkg)` packages (`r req_pkg_fmt`) for this chapter:

```{r}
#| label: cls-trees-installs
#| eval: false
#| echo: true

# fmt: skip
req_pkg <- 
  c("aorsf", "bonsai", "evtree", "partykit", "probably", "rpart",   "rules",
    "spatialsample", "tidymodels")

# Check to see if they are installed:
pkg_installed <- vapply(req_pkg, rlang::is_installed, logical(1))

# Install missing packages:
if (any(!pkg_installed)) {
  install_list <- names(pkg_installed)[!pkg_installed]
  pak::pak(install_list)
}
```

Let's load the meta package and manage some between-package function conflicts. 

```{r}
#| label: start-tidymodels
#| results: hide
#| message: false
#| warning: false

library(tidymodels)
library(spatialsample)
library(bonsai)
library(rules)
library(probably)
library(mirai)

tidymodels_prefer()
theme_set(theme_bw() + theme(legend.position = "top"))
daemons(parallel::detectCores())

# Some general settings for tuning classification models:
cls_mtr <- metric_set(brier_class, roc_auc, pr_auc, mn_log_loss)
ctrl_grid <-
  control_grid(
    save_pred = TRUE,
    save_workflow = TRUE,
    parallel_over = "everything"
  )
```

As before, we'll load the objects that were already computed for the forestation data: 

```{r}
#| label: data-load

"https://raw.githubusercontent.com/aml4td/website/main/RData/forested_data.RData" |>
  url() |>
  load()
```


## Specific Tree-Based Models {#sec-cls-tree-models}

### CART {#sec-cls-cart}


```{r}
#| label: rpart-fit
library(rpart)
cart_fit <- rpart(class ~ county + elevation, data = forested_train)
```


```{r}
#| label: rpart-rule
path.rpart(cart_fit, 11)
```

### C4.5 and C5.0 {#sec-cls-c50}



### Conditional Inference Trees {#sec-cls-cit}



### Globally Estimated Trees {#sec-cls-tree-global}



### Oblique Trees {#sec-cls-oblique}




## Rule-Based Models {#sec-cls-rules}



## The Instability of Trees {#sec-cls-tree-unstable}



## Comparisons to Previous Models {#sec-cls-tree-compare}
